Bow shocks, bow waves, and dust waves

### 2017-12-19 papers/Papers1234/sec-dust-bow-intro.tex
Added table of star parameters and reorganized a bit

We will mainly consider the canonical case of a bow around a star of
bolometric luminosity, \(L\), with a radiatively driven
wind,\footnote{In \S~\ref{sec:case-inside-out} we consider the case of
  an external luminous source.} which is immersed in an external
stream of gas and dust with density, \(\rho\), and velocity, \(v\).  The
size and shape of the bow is determined by a generalized balance of
pressure (or, equivalently, momentum) between internal and external
sources.  We assume that the stream is supersonic and super-alfvenic,
so that the external pressure is dominated by the ram pressure:
\(\rho v^2\). We first consider the case where the dust grains and gas
are perfectly coupled by collisions (although this assumption is
relaxed below in \S~\ref{sec:imperf-coupl-betw}).

The internal pressure is the sum of wind ram pressure and the
effective radiation pressure that acts on the bow shell.  The
radiative momentum loss rate of the star is \(L/c\) and the wind
momentum loss rate can be expressed as
\begin{equation}
  \label{eq:wind-efficiency}
  \dot{M} V = \eta L / c \ , 
\end{equation}
where \(\eta\) is the momentum efficiency of the wind, which is typically
\(< 1\) \citep{Lamers:1999b}. If the optical depth is very large, then
all of the stellar radiative momentum, emitted with rate \(L/c\), is
trapped by the bow shell.  In the single scattering
limit,\footnote{Why multiple scattering is not important.} and
temporarily neglecting the wind, then pressure balance at the bow
apex, a distance \(R_0\) along the symmetry axis from the star is given by
\begin{equation}
  \label{eq:rad-press-balance-thick}
  \frac{L}{4 \pi c R_0^2} = \rho v^2 \ ,
\end{equation}
which gives a fiducial bow shock radius in this optically thick limit
as
\begin{equation}
  \label{eq:Rstar}
  R_* = \left(\frac{L}{4\pi c \rho v^2}\right)^{1/2} \ .
\end{equation}

We now consider the opposite, optically thin limit.  Although dust
grains typically constitute only a small fraction
\(Z\grain \sim 0.01\) of the mass of the external stream, they
nevertheless dominate the broad-band opacity at FUV, optical and IR
wavelengths if they are present.\footnote{%
  At EUV wavelengths (\(\lambda < \SI{912}{\angstrom}\)), gas opacity
  dominates if the hydrogen neutral fraction is larger than
  \(\approx 0.001\), see discussion of ionization front trapping
  below.} %
If the total opacity (gas plus dust) per total mass (gas plus dust) is
\(\kappa\) (with units of \si{cm^2.g^{-1}}), then the radiative
acceleration in the optically thin limit is
\begin{equation}
  \label{eq:rad-accel}
  a_{\text{rad}} = \frac{\kappa L}{4 \pi c R^2} \ , 
\end{equation}
so that an incoming stream with initial velocity, \(v_\infty\), can be
brought to rest by radiation alone\footnote{%
  For simplicity, we here ignore the effects of pressure gradients and
  shocks, which are important as the velocity approaches the sound
  speed, \(\sound\). In \S~XX below, we show that the resultant
  corrections to \(R_0\) are of order \(\sound / v_\infty\).} %
at a distance \(R_0\) where
\begin{equation}
  \label{eq:rad-poten}
  \int_{R_0}^\infty a_{\text{rad}} \, dr = \tfrac12 v^2 \ , 
\end{equation}
yielding
\begin{equation}
  \label{eq:rad:R0}
  R_0 = \frac{\kappa L}{2\pi c v_\infty^2} \ .
\end{equation}
On the other hand, we can also employ a similar argument to the
optically thick case above by approximating the bow shell as a
surface, and balancing stellar radiation pressure against the ram
pressure of the incoming stream.  The important difference when the
shell is not optically thick is that only a fraction \(1 - e^{-\tau}\) of
the radiative momentum is absorbed by the bow, so that equation~\eqref{eq:rad-press-balance-thick} is replaced with 
\begin{equation}
  \label{eq:rad-press-balance-tau}
  \frac{L (1 - e^{-\tau})}{4 \pi c R_0^2} = \rho v^2 \ .
\end{equation}
In the optically thin limit, \(1 - e^{-\tau} \approx \tau\), so these two
descriptions can be seen to agree so long as
\(\tau = 2 \kappa \rho R_0\), which we will assume to be generally valid.

@@

We now consider the application to bow shocks around main sequence OB
stars, expressing stellar and ambient parameters in terms of typical
values as follows:
\begin{gather*}
  \label{eq:stellar-parameters}
  \dot{M}_{-7} = \dot{M} / \bigl(\SI{e-7}{M_\odot.yr^{-1}}\bigr) \\
  V_3 = V / \bigl(\SI{1000}{km.s^{-1}}\bigr) \\
  L_4 = L / \bigl(\SI{e4}{L_\odot}\bigr) \\
  v_{10} = v_\infty / \bigl( \SI{10}{km.s^{-1}} \bigr) \\
  n = (\rho / \bar{m}) / \bigl( \SI{1}{cm^{-3}} \bigr) \\
  \kappa_{600} = \kappa / \bigl( \SI{600}{cm^2.g^{-1}} \bigr) \ ,
\end{gather*}
where \(\bar{m}\) is the mean mass per hydrogen nucleon
(\(\bar{m} \approx 1.3 m_{\text{p}} \approx \SI{2.17e-24}{g}\) for solar
abundances).  Note that \(\kappa = \SI{600}{cm^2.g^{-1}}\) corresponds to a
cross section of \(\approx \SI{e-21}{cm^2}\) per hydrogen nucleon, which is
typical for interstellar medium dust \citep{Bertoldi:1996a} at far
ultraviolet wavelengths, where OB stars emit most of their radiation.
In terms of these parameters, we can express the stellar wind momentum
efficiency as
\begin{equation}
  \label{eq:wind-eta-typical}
  \eta = \num{0.495} \,\dot{M}_{-7} \,V_3  \,L_4^{-1}
\end{equation}
and the fiducial radius and optical depth as
\begin{align}
  \label{eq:fiducial-typical}
  R_* / \si{pc} &= \num{2.21} \, (L_4 / n)^{1/2} \,v_{10}^{-1} \\
  \tau_* &= \num{0.0089} \,\kappa_{600} \, (L_4 \,n)^{1/2} \,v_{10}^{-1}
\end{align}
\subsection{Imperfect coupling between gas and dust}
\label{sec:imperf-coupl-betw}

>>>>

In this section, we investigate the different types of bow interaction
that will occur in different regions of parameter space. We will
mainly treat the canonical case of a bow around a star of bolometric
luminosity, \(L\), with a radiatively driven wind,\footnote{%
  In \S~\ref{sec:case-inside-out} we consider the case of an external
  luminous source.} %
which is immersed in an external stream of gas and dust with density,
\(\rho\), and velocity, \(v\).  The size and shape of the bow is
determined by a generalized balance of pressure (or, equivalently,
momentum) between internal and external sources.  We assume that the
stream is supersonic and super-alfvenic, so that the external pressure
is dominated by the ram pressure: \(\rho v^2\).

\subsection{Strong gas-grain coupling}
\label{sec:strong-gas-grain}

We first consider the case where the dust grains and gas are perfectly
coupled by collisions.\footnote{%
  Cases where this assumption does not hold are investigated below in
  \S~\ref{sec:imperf-coupl-betw}.} %
Although dust grains typically constitute only a small fraction
\(Z\grain \sim 0.01\) of the mass of the external stream, they
nevertheless dominate the broad-band opacity at FUV, optical and IR
wavelengths if they are present.\footnote{%
  At EUV wavelengths (\(\lambda < \SI{912}{\angstrom}\)), gas opacity
  dominates if the hydrogen neutral fraction is larger than
  \(\approx 0.001\), see discussion of ionization front trapping
  below.} %
The strong coupling assumption means that all the radiative forces
applied to the dust grains are directly felt by the gas also.

The internal pressure is the sum of wind ram pressure and the
effective radiation pressure that acts on the bow shell.  The
radiative momentum loss rate of the star is \(L/c\) and the wind
momentum loss rate can be expressed as
\begin{equation}
  \label{eq:wind-efficiency}
  \dot{M} V = \eta L / c \ , 
\end{equation}
where \(\eta\) is the momentum efficiency of the wind, which is typically
\(< 1\) \citep{Lamers:1999b}. If the optical depth is very large, then
all of the stellar radiative momentum, emitted with rate \(L/c\), is
trapped by the bow shell.  In the single scattering limit,\footnote{%
  Although it may seem inconsistent to assume single scattering in the
  case of high optical depths, this is defensible for the following
  reasons. (1)~The grain albedo is not that high (typically
  \(\sim 0.5\) at ultraviolet through optical wavelengths). (2)~The
  scattered radiation field is more isotropic than the stellar field,
  leading to cancellation in the radiative
  flux. (3)~Absorbed radiation is re-emitted at infrared
  wavelengths, where the dust opacity is very much lower.} %
and temporarily neglecting the wind, then pressure balance at the bow
apex, a distance \(R_0\) along the symmetry axis from the star is
given by
\begin{equation}
  \label{eq:rad-press-balance-thick}
  \frac{L}{4 \pi c R_0^2} = \rho v^2 \ ,
\end{equation}
which yields a fiducial bow shock radius in this optically thick limit
as
\begin{equation}
  \label{eq:Rstar}
  R_* = \left(\frac{L}{4\pi c \rho v^2}\right)^{1/2} \ .
\end{equation}

We now consider the opposite, optically thin limit.  If the total
opacity (gas plus dust) per total mass (gas plus dust) is \(\kappa\) (with
units of \si{cm^2.g^{-1}}), then the radiative acceleration is
\begin{equation}
  \label{eq:rad-accel}
  a_{\text{rad}} = \frac{\kappa L}{4 \pi c R^2} \ , 
\end{equation}
so that an incoming stream with initial velocity, \(v_\infty\), can be
brought to rest by radiation alone\footnote{%
  For simplicity, we here ignore the effects of pressure gradients and
  shocks, which are important as the velocity approaches the sound
  speed, \(\sound\). In \S~XX below, we show that the resultant
  corrections to \(R_0\) are of order \(\sound / v_\infty\).} %
at a distance \(R_0\) where
\begin{equation}
  \label{eq:rad-poten}
  \int_{R_0}^\infty a_{\text{rad}} \, dr = \tfrac12 v^2 \ , 
\end{equation}
yielding
\begin{equation}
  \label{eq:rad:R0}
  R_0 = \frac{\kappa L}{2\pi c v_\infty^2} \ .
\end{equation}
On the other hand, we can also employ a similar argument to the
optically thick case above by approximating the bow shell as a
surface, and balancing stellar radiation pressure against the ram
pressure of the incoming stream.  The important difference when the
shell is not optically thick is that only a fraction \(1 - e^{-\tau}\) of
the radiative momentum is absorbed by the bow, so that
equation~\eqref{eq:rad-press-balance-thick} is replaced with
\begin{equation}
  \label{eq:rad-press-balance-tau}
  \frac{L (1 - e^{-\tau})}{4 \pi c R_0^2} = \rho v^2 \ .
\end{equation}
In the optically thin limit, \(1 - e^{-\tau} \approx \tau\), so these two
descriptions can be seen to agree so long as
\(\tau = 2 \kappa \rho R_0\), which we will assume to hold generally.

@@

\begin{table*}
  \centering
  \caption{Stellar parameters for main-sequence OB stars}
  \label{tab:stars}
  \begin{tabular}{l S S S S S S S S l}
    \toprule
    & {\(M / \si{M_\odot}\)} & {\(L_4\)}
    & {\(\dot{M}_{-7}\)} & {\(V_3\)} & {\( \eta \)}
    & {\(T_{\text{eff}} / \si{kK}\)} & {\(\lambda_{\text{eff}}\) / \si{\um}}
    & {\(S_{49}\)} & Figures 
    \\
    \midrule
    & 10 & 0.63 & 0.003 & 1.082 & 0.0026 & 25.2 & 0.115 & e-4
                   & \ref{fig:zones-v-n-plane}a,
                     \ref{fig:decouple-v-n-plane},
                     \ref{fig:decouple-v40-versus-n} \\
    Main-sequence OB stars
    & 20 & 5.5 & 0.42 & 1.167 & 0.0446 & 33.9 & 0.086 & 0.1
                   & \ref{fig:zones-v-n-plane}b\\
    & 40 & 22 & 5.1 & 1.451 & 0.1682 & 42.5 & 0.068 & 1.0
                   & \ref{fig:zones-v-n-plane}c\\[\smallskipamount]
    Red supergiant star
    & 20 & 15.6 & 100 & 0.015 & 0.0476 & 4 & 0.725 & 0
                   & \\ 
    \bottomrule
  \end{tabular}
\end{table*}

We now consider the application to bow shocks around main sequence OB
stars, expressing stellar and ambient parameters in terms of typical
values as follows:
\begin{gather*}
  \label{eq:stellar-parameters}
  \dot{M}_{-7} = \dot{M} / \bigl(\SI{e-7}{M_\odot.yr^{-1}}\bigr) \\
  V_3 = V / \bigl(\SI{1000}{km.s^{-1}}\bigr) \\
  L_4 = L / \bigl(\SI{e4}{L_\odot}\bigr) \\
  v_{10} = v_\infty / \bigl( \SI{10}{km.s^{-1}} \bigr) \\
  n = (\rho / \bar{m}) / \bigl( \SI{1}{cm^{-3}} \bigr) \\
  \kappa_{600} = \kappa / \bigl( \SI{600}{cm^2.g^{-1}} \bigr) \ ,
\end{gather*}
where \(\bar{m}\) is the mean mass per hydrogen nucleon
(\(\bar{m} \approx 1.3 m_{\text{p}} \approx \SI{2.17e-24}{g}\) for solar
abundances).  Note that \(\kappa = \SI{600}{cm^2.g^{-1}}\) corresponds to a
cross section of \(\approx \SI{e-21}{cm^2}\) per hydrogen nucleon, which is
typical for interstellar medium dust \citep{Bertoldi:1996a} at far
ultraviolet wavelengths, where OB stars emit most of their radiation.
In terms of these parameters, we can express the stellar wind momentum
efficiency as
\begin{equation}
  \label{eq:wind-eta-typical}
  \eta = \num{0.495} \,\dot{M}_{-7} \,V_3  \,L_4^{-1}
\end{equation}
and the fiducial radius and optical depth as
\begin{align}
  \label{eq:fiducial-typical}
  R_* / \si{pc} &= \num{2.21} \, (L_4 / n)^{1/2} \,v_{10}^{-1} \\
  \tau_* &= \num{0.0089} \,\kappa_{600} \, (L_4 \,n)^{1/2} \,v_{10}^{-1} \ .
\end{align}
In Figure~\ref{fig:zones-v-n-plane}, we show results for the bow size
(apex distance, \(R_0\)) as a function of the density, \(n\), and
relative velocity, \(v_\infty\), of the external stream, with each panel
corresponding to a particular star, with parameters as shown in
Table~\ref{tab:stars}.

\subsection{Imperfect coupling between gas and dust}
\label{sec:imperf-coupl-betw}
